---
title: "3302 Final Project Doctor Visits"
author: "Charles Costanzo"
date: "2023-04-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/charlescostanzo/College/Sp 2023/Stat 3302/Final Project/")

library(tidyverse)
library(AER)
library(MASS)
library(stargazer)
library(pscl)
library(kableExtra)
data("DoctorVisits")
doctor <- DoctorVisits
```

Create a variable table
```{r}
variable_table <- data.frame(Variable = c("visits","gender","age","income","illness","reduced",
                        "health","private","freepoor","freerepat","nchronic",
                        "lchronic"),
           Description = c("Number of doctor visits in past 2 weeks.",
                           "Factor indicating gender.",
                           "Age in years divided by 100.",
                           "Annual income in tens of thousands of dollars.",
                           "Number of illnesses in past 2 weeks.",
                           "Number of days of reduced activity in past 2 weeks
                           due to illness or injury.",
                           "General health questionnaire score using Goldberg's method.",
                           "Factor. Does the individual have private health insurance?",
                           "Factor. Does the individual have free government 
                           health insurance due to low income?",
                           "Factor. Does the individual have free government 
                           health insurance due to old age, disability or veteran status?",
                           "Factor. Is there a chronic condition not limiting activity?",
                           "Factor. Is there a chronic condition limiting activity?"))

# Create table for export
knitr::kable(variable_table) %>%
  kable_styling(full_width = F,
                font_size = 12,
                position = "left")
```

\newpage

Create density plot for visits
```{r}
library(ggthemes)
p <- ggplot(data = doctor) +
  geom_density(aes(doctor$visits)) + 
  xlab("Number of Doctor Visits in Past 2 Weeks") + 
  ylab("Density") +
  ggtitle("Density of Dependent Variable") +
  scale_x_continuous(labels = c(0, 1,2,3,4,5,6,7,8,9,10),
                     breaks = c(0, 1,2,3,4,5,6,7,8,9,10),
                     lim = c(-1,10)) + 
  theme_hc() + scale_colour_hc()

# theme(rect = element_rect(fill = "transparent"),
#         panel.background = element_rect(fill = "transparent",
#         colour = NA_character_))
```

```{r}
poisson_y <- doctor %>%
  group_by(visits) %>%
  summarize(count = n(),
            prob = count/5190)

x_vec <- poisson_y$visits # vector of sample elements
y_vec <- poisson_y$prob # probability that count = x

i <- 0
plot(x_vec, y_vec, ylim = c(0, 1), pch = 19, 
     xlab = "Number of Doctor Visits in the Past Two Weeks (x)",
     ylab = "Probability that Number of Visits = x", xaxt = "n",
     main = "Distribution of Visits") +
  axis(side = 1, at = x_vec, tick = TRUE)
for (i in 1:length(x_vec)) { # connect dots to x axis
  segments(x_vec[i], 0, x_vec[i], y_vec[i], lty = 2)
}
```


Save plot as a .png file in "plots" folder
```{r}
ggsave("visits_density.png", p, bg="transparent",
       path = paste(getwd(),"/plots",sep=""))
```

\newpage

Create distribution plot for `private` vs `insurance`

```{r}
p2 <- ggplot(data = doctor) +
  geom_violin(aes(x = freepoor,
                  y = income,
                  fill = as.factor(freepoor))) +
  geom_boxplot(aes(x = freepoor,
                   y = income),
               width = 0.1) + 
  xlab("") + 
  ylab("Annual Income (Tens of Thousands of Australian Dollars)") +
  ggtitle("Parallel Box and Violin Plots for Income by Free Low Income Insurance") +
  theme_hc() + scale_fill_manual(name = "Free Government Insurance due to Low Income",
    values = c("#6B9080","#A4C3B2")) + 
  coord_flip()
```

Save plot as a .png file in "plots" folder
```{r}
ggsave("private_income_boxplot.png", p2, bg="transparent",
       path = paste(getwd(),"/plots",sep=""))
```

\newpage

Summarize `visits` values
```{r}
doctor %>%
  group_by(as.factor(visits)) %>%
  summarize(count = n())
```

\newpage

Get mean number of doctor visits by insurance type
```{r}
mean_visits_insur <- doctor %>%
  group_by(private, freepoor, freerepat) %>%
  summarize(mean_visits = mean(visits, na.rm = TRUE))
mean_visits_insur

mean_visits_insur_table <- data.frame(
  Insurance = c("Private","Free Government Insurance due to Low Income",
      "Free Government Insurance due to Old Age, Disability, or Veteran Status",
      "No Insurance"),
  `Mean Doctor Visits` = 
    c(mean_visits_insur$mean_visits[mean_visits_insur$private == "yes"],
      mean_visits_insur$mean_visits[mean_visits_insur$freepoor == "yes"],
      mean_visits_insur$mean_visits[mean_visits_insur$freerepat == "yes"],
mean_visits_insur$mean_visits[mean_visits_insur$private == "no" & mean_visits_insur$freepoor == "no" & mean_visits_insur$freerepat == "no"]
    ))

mean_visits_insur_table$Mean.Doctor.Visits <-
  round(mean_visits_insur_table$Mean.Doctor.Visits, 3)

names(mean_visits_insur_table) <- c("Insurance Type", "Mean Doctor Visits (past 2 weeks)")

knitr::kable(mean_visits_insur_table) %>%
  kable_styling(full_width = F,
                font_size = 16,
                position = "left")
```

\newpage

Get mean number of doctor visits by gender
```{r}
mean_visits_gender <- doctor %>%
  group_by(gender) %>%
  summarize(Mean.Doctor.Visits = mean(visits, na.rm = TRUE))

mean_visits_gender_table <- data.frame(
  Gender = c("Male","Female"),
  Mean.Doctor.Visits = 
    c(mean_visits_gender$Mean.Doctor.Visits[mean_visits_gender$gender == "male"],
      mean_visits_gender$Mean.Doctor.Visits[mean_visits_gender$gender == "female"]))

names(mean_visits_gender_table) <- c("Sex", "Mean Doctor Visits (past 2 weeks)")

knitr::kable(mean_visits_gender_table) %>%
  kable_styling(full_width = F,
                font_size = 30,
                position = "center")
```

\newpage

# Model Description

Let $\textbf{Y}$ be our vector of data containing the number of doctor visits within the past two weeks, `visits`.

$$\text{Assume } \{Y_1\ldots,Y_{n=5190}\} \overset{ind}{\sim} Poisson(\lambda_i)$$

We will use the log link.

Poisson GLM:

$$\eta_i = g(\lambda_i) = log(\lambda_i) = \textbf{x}_i^{\intercal}\boldsymbol{\beta}, \; \; i = 1,\ldots,n=5190$$

where 

 - $\eta_i$ is the estimated log mean number of doctor visits in the past two weeks for individual $i$ (according to their specific covariate values $\textbf{x}_i$)
 
 - $\textbf{x}_i=(x_{i,1},\ldots,x_{i,p=12})^\intercal$ are the $p$ covariates (including an intercept) for individual $i$
 
 - $\boldsymbol{\beta} = (\beta_0,\beta_1,\ldots,\beta_{p=12})$ is our unknown coefficient vector. 
 
 \newpage
 
run first Poisson model for visits
```{r}
model1 <- glm(visits ~ .,
              family = "poisson",
              data = doctor)
summary(model1)
```

\newpage

Create Analysis of Deviance Table
```{r}
anova(model1, test = "Chisq")
```

\newpage

Conduct a Chi-Square Test at the $\alpha = 0.05$ level to demonstrate that the `freerepat`, `nchronic`, and `lchronic` variables should be removed from the model.


$H_0:$ Reduced Model (without `freerepat`, `nchronic`, and `lchronic`) is good enough vs.

$H_1:$ Full Model (including all predictor variables) is Needed


 - $D_{full} = 4380.1$
 
 - $D_{reduced} = 4385.5$
 
*Test Statistic:*

$$\Delta D = D_{reduced} - D_{full} = 4385.5 - 4380.1$$
$$\Rightarrow \Delta D = 5.4$$

Under $H_0, \Delta D \sim \chi^{2}_{pFull-pReduced = 12-9}$ 
$$\Rightarrow \Delta D \sim\chi^2_3$$
**Rejection Rule:** Reject if $\Delta D >$ `qchisq(1-0.05,3)` = 7.814728

Since our Test Statistic $\Delta D = 5.4 < 7.8,$ we fail to reject the null hypothesis at the $\alpha = 0.05$ significance level.

**p value:** `pchisq(5.4, 3, lower.tail = FALSE)` = 0.1447436

**Conclusion:**

The reduced model is good enough (that does not include `freerepat`, `nchronic`, and `lchronic`) and provides a better fit to the data set than the full model that includes those three covariates. We will now drop these covariates and fit a reduced model.

\newpage

```{r}
model1_reduced <- glm(visits ~ gender + age + income + illness + reduced + health + 
    private + freepoor,
    family = "poisson",
    data = doctor)
summary(model1_reduced)
```

\newpage

Check for overdispersion
``` {r}
var(doctor$visits)
mean(doctor$visits)

dispersiontest(model1, alternative = "greater")
```

There is indeed overdispersion, so run a quasipoisson model

\newpage

Run quasipoisson model
``` {r}
model2_reduced <- glm(visits ~ gender + age + income + illness + reduced + health + 
    private + freepoor,
    family = "quasipoisson",
    data = doctor)
summary(model2_reduced)
```

\newpage

Also try a zero-inflated Poisson since there are a lot of zeros (see density plot)
```{r}
model3 <- zeroinfl(visits ~ .,
    data = doctor)
summary(model3)
```

\newpage

Run a model to see what variables impact whether or not someone has private insurance
``` {r}
model4 <- glm(private ~ .,
              family = "binomial",
              data = doctor)
summary(model4)
```

\newpage

Create stargazer tables for models

```{r}
# stargazer(model1,model1_reduced,model2_reduced, type = "html", out = "count_models.html")

# stargazer(model4, type = "html", out = "logistic_model.html")
```

\newpage

Create a function to run diagnostics on the count models
```{r}
poisson_diagnostics <- function (model){
  fits <- fitted(model)
  deviance.resids <- resid(model1) ## deviance is the default
  par(mfrow = c(2,2), cex = 0.3)
  plot(fits, deviance.resids); abline(h = 0)
  qqnorm(deviance.resids); qqline(deviance.resids)

  pearson.resids <- resid(model, type = "pearson")
  plot(fits, pearson.resids);abline(h = 0)
  qqnorm(pearson.resids) ; qqline(pearson.resids)
}
```

\newpage

```{r}
poisson_diagnostics(model1)
```

\newpage

```{r}
poisson_diagnostics(model1_reduced)
```

\newpage

```{r}
poisson_diagnostics(model2_reduced)
```

\newpage

```{r}
poisson_diagnostics(model3)
```

\newpage

# Count models for Doctor Visits

\begin{table}[!htbp] \centering 
  \caption{} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textit{Dependent variable:}} \\ 
\cline{2-4} 
\\[-1.8ex] & \multicolumn{3}{c}{visits} \\ 
\\[-1.8ex] & \textit{Poisson} & \textit{negative} & \textit{zero-inflated} \\ 
 & \textit{} & \textit{binomial} & \textit{count data} \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 genderfemale & 0.156$^{***}$ & 0.216$^{***}$ & $-$0.020 \\ 
  & (0.056) & (0.070) & (0.072) \\ 
  & & & \\ 
 age & 0.279$^{*}$ & 0.331 & $-$0.002 \\ 
  & (0.166) & (0.208) & (0.218) \\ 
  & & & \\ 
 income & $-$0.187$^{**}$ & $-$0.156 & $-$0.214$^{*}$ \\ 
  & (0.085) & (0.104) & (0.110) \\ 
  & & & \\ 
 illness & 0.186$^{***}$ & 0.215$^{***}$ & 0.044$^{*}$ \\ 
  & (0.018) & (0.024) & (0.025) \\ 
  & & & \\ 
 reduced & 0.127$^{***}$ & 0.144$^{***}$ & 0.083$^{***}$ \\ 
  & (0.005) & (0.007) & (0.006) \\ 
  & & & \\ 
 health & 0.031$^{***}$ & 0.038$^{***}$ & 0.023$^{**}$ \\ 
  & (0.010) & (0.014) & (0.011) \\ 
  & & & \\ 
 privateyes & 0.126$^{*}$ & 0.116 & $-$0.022 \\ 
  & (0.072) & (0.086) & (0.097) \\ 
  & & & \\ 
 freepooryes & $-$0.438$^{**}$ & $-$0.497$^{**}$ & $-$0.382 \\ 
  & (0.180) & (0.211) & (0.241) \\ 
  & & & \\ 
 freerepatyes & 0.084 & 0.146 & $-$0.212$^{*}$ \\ 
  & (0.092) & (0.116) & (0.119) \\ 
  & & & \\ 
 nchronicyes & 0.117$^{*}$ & 0.098 & $-$0.012 \\ 
  & (0.067) & (0.079) & (0.093) \\ 
  & & & \\ 
 lchronicyes & 0.151$^{*}$ & 0.183$^{*}$ & 0.002 \\ 
  & (0.082) & (0.103) & (0.102) \\ 
  & & & \\ 
 Constant & $-$2.098$^{***}$ & $-$2.276$^{***}$ & $-$0.550$^{***}$ \\ 
  & (0.102) & (0.123) & (0.144) \\ 
  & & & \\ 
\hline \\[-1.8ex] 
Observations & 5,190 & 5,190 & 5,190 \\ 
Log Likelihood & $-$3,355.850 & $-$3,199.838 & $-$3,180.927 \\ 
$\theta$ &  & 0.930$^{***}$  (0.087) &  \\ 
Akaike Inf. Crit. & 6,735.701 & 6,423.676 &  \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 

\newpage

# Logistic Regression Model for Private Insurance

\begin{table}[!htbp] \centering 
  \caption{} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{1}{c}{\textit{Dependent variable:}} \\ 
\cline{2-2} 
\\[-1.8ex] & private \\ 
\hline \\[-1.8ex] 
 visits & 0.074 \\ 
  & (0.057) \\ 
  & \\ 
 genderfemale & 0.818${***}$ \\ 
  & (0.074) \\ 
  & \\ 
 age & 3.726$^{***}$ \\ 
  & (0.242) \\ 
  & \\ 
 income & 0.689$^{***}$ \\ 
  & (0.101) \\ 
  & \\ 
 illness & 0.025 \\ 
  & (0.032) \\ 
  & \\ 
 reduced & $-$0.027$^{*}$ \\ 
  & (0.016) \\ 
  & \\ 
 health & $-$0.022 \\ 
  & (0.019) \\ 
  & \\ 
 freepooryes & $-$18.413 \\ 
  & (418.751) \\ 
  & \\ 
 freerepatyes & $-$20.264 \\ 
  & (190.618) \\ 
  & \\ 
 nchronicyes & 0.241$^{***}$ \\ 
  & (0.082) \\ 
  & \\ 
 lchronicyes & 0.285$^{**}$ \\ 
  & (0.139) \\ 
  & \\ 
 Constant & $-$1.805$^{***}$ \\ 
  & (0.119) \\ 
  & \\ 
\hline \\[-1.8ex] 
Observations & 5,190 \\ 
Log Likelihood & $-$2,330.324 \\ 
Akaike Inf. Crit. & 4,684.648 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 